### 2016-03-19 doc/thesis/conclusion.tex
Draft of contact network section done

@@

It is important to note that our kernel-ABC method takes a transmission tree as
input, rather than a viral phylogeny. Thus, we have left the estimation of a
transmission tree up to the user. There were two reasons for this choice. First
and foremost, we wished to avoid the extra complexity and keep the number of
estimated parameters small. In theory, it is possible to incorporate the
process by which a viral phylogeny is generated along with a transmission tree
into our method, for example by simulating within-host dynamics. Although this
may be an avenue for future extension, we felt that it would obscure the
primary purpose of this work, which is to study contact network parameters.
Second, there are a number of different methods available for inferring
transmission trees~\autocite{didelot2014bayesian, yima2012unravelling,
jombart2011reconstructing, cottam2008integrating, poon2015phylodynamic}, some
of which incorporate geographic and/or epidemiological data not accomodated by
our method. We therefore felt it would be best to allow researchers to use
their own preferred tree building method.

>>>>

The two-step process of simulating a contact network and subsequently allowing
an epidemic to spread over that network carries with it the assumption that the
contact network is static over the duration of the epidemic. Clearly this
assumption is invalid, as people make and break partnerships on a regular
basis. Our work has not addressed this assumption, primarily due to our desire
to avoid the additional complexity required to address the dynamic nature of
networks. This simplifying assumption is made by most studies using contact
network models in an epidemiological context~\autocite{welch2011statistical,
bansal2007individual}. However, in principle, kernel-\gls{ABC} could be
adapted to dynamic contact networks by using a method such as that developed by
\textcite{robinson2012dynamics} to simulate a dynamic contact network, while
concurrently simulating the spread of an epidemic.

@@

It is important to note that our kernel-ABC method takes a transmission tree as
input, rather than a viral phylogeny. Thus, we have left the estimation of a
transmission tree up to the user. There were two reasons for this choice. First
and foremost, we wished again to avoid extra complexity and keep the number of
estimated parameters small. In theory, it is possible to incorporate the
process by which a viral phylogeny is generated along with a transmission tree
into our method, for example by simulating within-host dynamics. Although this
may be an avenue for future extension, we felt that it would obscure the
primary purpose of this work, which is to study contact network parameters.
Second, there are a number of different methods available for inferring
transmission trees~\autocite{didelot2014bayesian, yima2012unravelling,
jombart2011reconstructing, cottam2008integrating, poon2015phylodynamic}, some
of which incorporate geographic and/or epidemiological data not accomodated by
our method. We therefore felt it would be best to allow researchers to use
their own preferred tree building method.

### 2016-03-19 doc/thesis/intro.tex
Draft of contact network section done

Epidemics spread through populations of hosts through \defn{contacts} between
those hosts. The definition of contact depends on the mode of transmission of
the pathogen in question. For an airborne pathogen like influenza, a contact
may be simple physical proximity, while for \gls{HIV}, contact could be via
unprotected sexual relations or blood-to-blood contact (such as through needle
sharing). A \defn{contact network} is a graphical representation of a host
population and the contacts among its members~\autocite{klovdahl1985social,
morris1993epidemiology, keeling2005networks}. The \defn{nodes} in the network
represent hosts, and \defn{edges} or \defn{links} represent contacts between
them. A contact network is shown in \cref{fig:contactnet} (left).

Edges in a contact networks may be \defn{directed}, representing one-way
transmission risk, or \defn{undirected}, representing symmetric transmission
risk. For example, a network for an airborne epidemic would use undirected
edges, because the same physical proximity is required for a host to infect or
to become infected. However, an infection which may be spread through
blood-to-blood contact through transfusions transfusions would use undirected
edges, since the donor has no chance of transmitting to the recipient. Directed
edges are also useful when the transmission risk is not equal between the
hosts, such as with HIV transmission, where acting as the receptive partner
carries a higher risk of infection than acting as the insertive partner. In
this case, a contact could be represented by two directed edges, one in each
direction between the two hosts, with the edges annotated by what kind of risk
they imply~\autocite{wasserman1994social}. An undirected contact network is
equivalent to a directed network where each contact is represented by two
symmetric directed edges. The \defn{degree} of a node in the network is how
many contacts it has. In directed networks, we may make the distinction between
\defn{out-degree} and \defn{in-degree}, which count respectively the number
incoming and outgoing edges. The set of edges attached to a node are referred
to as its \defn{incident} edges.

The path an epidemic takes through a contact network defines the
\defn{transmission network}, which is the subgraph of the contact network
consisting of infected nodes and the edges which led to their
infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). The
transmission network determines the topology of the transmission tree relating
the infected hosts. The initially infected node who introduces the epidemic
into the network, also known as the \defn{index case}, becomes the root of the
tree. Each time a transmission occurs, the lineage corresponding to the donor
host in the tree splits into two, representing the recipient lineage and the
continuation of the donor lineage. \Cref{fig:contactnet} illustrates this
correspondence. It's important to note that, although the order and timing of
transmissions determines the tree topology uniquely, the converse does not
hold. That is, for any given topology, there are in general many transmission
networks which would lead to that topology. In other words, it impossible to
distinguish who transmitted to whom from a transmission tree
alone~\autocite{bernard2007hiv}.

From a public health perspective, knowledge of contact networks has the
potential to be extremely useful. On a population level, network structure can
dramatically affect the speed and pattern of epidemic
spread~\autocite[\eg][]{barthelemy2005dynamical}. For example, epidemics are
expected to spread more rapidly in networks having the ``small world''
property, where the average path length between two nodes in the network is
relatively low~\autocite{watts1998collective}. Some sexually transmitted
infections would not be expected to survive in a homogeneously mixed
population, but their long-term persistance can be explained by contact
heterogeneity~\autocite{anderson1992infectious, pastor2001epidemic}. Hence, the
contact network can provide an idea of what to expect as an epidemic unfolds.
In terms of actionable information, vaccination strategies which would
irradicate an epidemic in a random network might not work if the network is
scale-free~\autocite[][see \cref{subsec:pa}]{keeling2005networks}. On a local
level, contact networks can be informative about the groups or individuals who
are at highest risk of acquiring or transmitting infection, and would therefore
benefit most from public health interventions~\autocite{wang2015targeting,
little2014using}.

Contact networks are a challenging type of data to collect, requiring extensive
epidemiological investigation in the form of contact
tracing~\autocite{morris1993epidemiology, welch2011statistical,
keeling2005networks}. Several less resource-intensive alternatives have been
explored. Some work focuses on the phylogenetic network, detecting patterns
such as clustering. However, the phylogenetic network may show little to no
agreement with a contact data obtained through epidemiological
methods~\autocite{yirrell1998molecular, resik2007limitations}. Some
studies~\autocite{brown2011transmission} have focused on parameters of the
transmission network, rather than the more complete contact network, because it
is possible to derive an estimate of the transmission network phylogenetically.

Rather than considering entire social networks, some epidemiological research
has focused on properties which can be investigated without extensive contact
tracing. The most common of these properties is the \defn{degree distribution}
of the network, which denotes the probability that a node has any given number
of links. Degree distributions for sexual networks can be estimated by simply
asking each person how many contacts they had in, say, the last
year~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk}.

There have been a few statistical methods developed to infer contact network
parameters without extensive contact tracing. The pioneering study in this
group was by \textcite{britton2002bayesian}, who developed a Bayesian method to
infer the $p$ parameter of an \gls{ER} network, along with the transmission and
removal rate parameters of the \gls{SI} model. Their method is able to use
either observed infection and removal times, or removal times only. However, it
is designed for only a small number of observations (fifteen and 42 cases in
their applications), and their estimates of $p$ for real outbreaks were mostly
uninformative (95\% confidence intervals [0.11-0.96] and [0.055-0.96]; $p$ is
bounded in [0, 1]). \textcite{groendyke2011bayesian} significantly updated and
extended the methodology of \citeauthor{britton2002bayesian}, and applied it to
a measles outbreak affecting 188 individuals. They were able to obtain a much
more informative estimate of $p$, although this data set included both symptom
onset and recovery times for all individuals. They also assumed that the entire
network was infected, so that the network size was exactly 188. It is unclear
how well their method would scale to networks with thousands of nodes.

@@

A \defn{scale-free} network is one whose degree distribution follows a power
law, meaning that the number of nodes in the network with degree $k$ is
proportional to $k^{-\gamma}$ for some constant
$\gamma$~\autocite{barabasi1999emergence}. Scale-free networks are
characterized by a large number of nodes of low degree, with relatively few
``hub'' nodes of very high degree.

Epidemiological surveys have indicated that human sexual networks tend to be
scale-free~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk}.
In this context, the high-degree hub nodes are known as
\defn{superspreaders}~\autocite{kemper1980identification}.

@@

There has been some contention of the idea that sexual networks are scale-free.
\textcite{handcock2004likelihood} fit several stochastic models of partner
formation to empirical degree distributions derived from population surveys of
sexual behaviour. They found that a negative binomial distribution, rather than
a power law, was the best fit to five out of six datasets, although the
difference in goodness of fit was extremely small in four out of these five.

@@

Clearly, the shape of a transmission tree is shaped by the underlying contact
network (\cref{fig:contactnet}). A number of studies have made progress in
quantifying this relationship. \textcite{o2010contact} simulated epidemics over
networks with four types of degree distribution. They then estimated the
Bayesian skyride~\autocite{minin2008smooth} population size trajectory in two
ways: from the phylogeny, using \gls{MCMC}; and from the incidence and
prevalence trajectories, using the method developed by Volz
\etal~\autocite{volz2009phylodynamics}. They found that the concordance between
the two skyrides, as well as the relationship between the skyride and
prevalence curve, was qualitatively different for each degree distribution.
\citeauthor{leventhal2012inferring} simulated transmission trees over \gls{ER},
\gls{WS}, \gls{BA}, and full networks with fixed number of nodes and mean
degree. They calculated Sackin's index of the simulated trees while varying the
epidemic, network, and sampling parameters, and found that the relationship
between these parameters and Sackin's index varies considerably among the
different network models. In summary, studies in this group have demonstrated
that network structure profoundly influences tree shape, but have not attempted
to quantitatively infer network parameters from observed trees.

>>>>

Epidemics spread through populations of hosts through \defn{contacts} between
those hosts. The definition of contact depends on the mode of transmission of
the pathogen in question. For an airborne pathogen like influenza, a contact
may be simple physical proximity, while for \gls{HIV}, contact could be via
unprotected sexual relations or blood-to-blood contact (such as through needle
sharing). A \defn{contact network} is a graphical representation of a host
population and the contacts among its members~\autocite{klovdahl1985social,
morris1993epidemiology, keeling2005networks}. The \defn{nodes} in the network
represent hosts, and \defn{edges} or \defn{links} represent contacts between
them. A contact network is shown in \cref{fig:contactnet} (left). Contact
networks are a particular type of \defn{social
network}~\autocite{moreno1953shall, barnes1954class}, which is a network in
which edges may represent any kind of social or economic relationship. Social
networks are frequently used in the social sciences to study phenomena where
relationships between people or entities are important \autocite[for a review
see][]{wasserman1994social}.

Edges in a contact networks may be \defn{directed}, representing one-way
transmission risk, or \defn{undirected}, representing symmetric transmission
risk. For example, a network for an airborne epidemic would use undirected
edges, because the same physical proximity is required for a host to infect or
to become infected. However, an infection which may be spread through
blood-to-blood contact through transfusions transfusions would use directed
edges, since the donor has no chance of transmitting to the recipient. Directed
edges are also useful when the transmission risk is not equal between the
hosts, such as with HIV transmission, where acting as the receptive partner
carries a higher risk of infection than acting as the insertive partner. In
this case, a contact could be represented by two directed edges, one in each
direction between the two hosts, with the edges annotated by what kind of risk
they imply~\autocite{wasserman1994social}. An undirected contact network is
equivalent to a directed network where each contact is represented by two
symmetric directed edges. The \defn{degree} of a node in the network is how
many contacts it has. In directed networks, we may make the distinction between
\defn{out-degree} and \defn{in-degree}, which count respectively the number
incoming and outgoing edges. The \defn{degree distribution} of a network
denotes the probability that a node has any given number of links. The set of
edges attached to a node are referred to as its \defn{incident} edges.

Epidemiological models most often assume some form of contact homogeneity. The
simplest models, such as the \gls{SIR} model, assume a completely homogeneously
mixed population, where every pair of contacts is equally likely. More
sophisticated models partition the population into groups with different
contact rates between and among each group. However, these models still assume
that every possible contact between a member of group $i$ and a member of group
$j$ is equally likely. This assumption is clearly unrealistic for the majority
of human communities, and can lead to significant errors in predicted epidemic
trajectories when there is substantial heterogeneity
present~\autocite{bansal2007individual, volz2007susceptible}. Contact networks
provide a way to relax this assumption by representing individuals and their
contacts explicitly. It is important to note that, although panmixia is an
unrealistic modelling assumption, it has not proven a substantial hurdle to
epidemic modelling in practice~\autocite{anderson1992infectious}. Using this
assumption, researchers have been able to derive estimates of the transmission
rate and the basic reproductive number of various outbreaks, which have agreed
with values obtained by on-the-ground data collection. Therefore, if one is
interested only in these population-level variables, the additional complexity
of contact network models may not be warranted. Rather, these models are most
useful when we are interested in properties of the network itself, such as
centrality, structural balance, and
transitivity~\autocite{wasserman1994social}.

From a public health perspective, knowledge of contact networks has the
potential to be extremely useful. On a population level, network structure can
dramatically affect the speed and pattern of epidemic
spread~\autocite[\eg][]{barthelemy2005dynamical, volz2008sir}. For example,
epidemics are expected to spread more rapidly in networks having the ``small
world'' property, where the average path length between two nodes in the
network is relatively low~\autocite{watts1998collective}. Some sexually
transmitted infections would not be expected to survive in a homogeneously
mixed population, but their long-term persistance can be explained by contact
heterogeneity~\autocite{anderson1992infectious, pastor2001epidemic}. Hence, the
contact network can provide an idea of what to expect as an epidemic unfolds.
In terms of actionable information, vaccination strategies which would
irradicate an epidemic in a random network might not work if the network is
scale-free~\autocite[][see \cref{subsec:pa}]{keeling2005networks}. On a local
level, contact networks can be informative about the groups or individuals who
are at highest risk of acquiring or transmitting infection, and would therefore
benefit most from public health interventions~\autocite{wang2015targeting,
little2014using}.

Contact networks are a challenging type of data to collect, requiring extensive
epidemiological investigation in the form of contact
tracing~\autocite{morris1993epidemiology, welch2011statistical,
keeling2005networks}. Therefore, it has been necessary to explore less
resource-intensive alternatives which still contain information about
population structure. For instance, it is possible to obtain limited
information about the contact network by individual interviews without contact
tracing. Variables which can be estimated in this fashion are referred to as
\defn{node-level} measures~\autocite{wasserman1994social}. One of the most
well-studied of these is the degree distribution, which can be estimated by
simply asking each person how many contacts they had in some interval of
time~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk}.

An alternative approach has been the analysis of other networks, which can be
estimated with phylogenetic methods from viral sequence data. Some work focuses
on the \defn{phylogenetic network}, in which two nodes are connected if the
genetic distance between their viral sequences is below some threshold.
Primarily, this work has focused on the detection of \defn{phylogenetic
clusters}, which are groups of individuals whose viral sequences are
significantly more similar to each other's than to the general population's.
The phylogenetic network is informative about ``hotspots'' of transmission and
can be used to identify demographic groups to whom targeted intervetions are
likely to have the greatest effect~\autocite{poon2014impact}.  However, this
network may show little to no agreement with a contact data obtained through
epidemiological methods~\autocite{yirrell1998molecular, resik2007limitations,
robinson2013dynamics}, and therefore may be a poor proxy for the contact
network. Other studies~\autocite{brown2011transmission} have investigated the
\defn{transmission network}, which is the subgraph of the contact network
consisting of infected nodes and the edges which led to their
infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). It is
possible to estimate the transmission network phylogenetically, although the
methods required for doing so are more sophisticated than for estimating the
phylogenetic network~\autocite{brown2011transmission}. These studies again
mostly focusing on clustering, and also on degree distributions.

Other statistical methods have been developed to infer contact network
parameters strictly from the timeline of an epidemic, using neither genetic
data nor reported contacts. \textcite{britton2002bayesian} developed a Bayesian
method to infer the $p$ parameter of an \gls{ER} network, along with the
transmission and removal rate parameters of the \gls{SI} model, using observed
infection and optionally removal times. However, it was designed for only a
small number of observations, and was unable to estimate $p$ independently from
the transmission rate.  \textcite{groendyke2011bayesian} significantly updated
and extended the methodology of \citeauthor{britton2002bayesian}, and applied
it to a measles outbreak affecting 188 individuals. They were able to obtain a
much more informative estimate of $p$, although this data set included both
symptom onset and recovery times for all individuals, and was unusual in that
the entire contact network was presumed to be infected. \textcite{volz2008sir}
developed differential equations describing the dynamics of the \gls{SIR} model
on a wide variety of random networks defined by their degree distributions.
Although the topic of estimation was not addressed in the original paper,
\citeauthor{volz2008sir}'s method could in principle be used to fit such models
to observed epidemic trajectories, similar to what is done with the ordinary
\gls{SIR} model. \textcite{volz2007susceptible} later extended the method to
dynamic contact networks and applied it to a sexual network relating 99
individuals investigated during a syphilis outbreak.

@@

A \defn{scale-free} network is one whose degree distribution follows a power
law, meaning that the number of nodes in the network with degree $k$ is
proportional to $k^{-\gamma}$ for some constant
$\gamma$~\autocite{barabasi1999emergence}. Scale-free networks are
characterized by a large number of nodes of low degree, with relatively few
``hub'' nodes of very high degree. Epidemiological surveys have indicated that
human sexual networks tend to be scale-free~\autocite{liljeros2001web,
schneeberger2004scale, colgate1989risk}. Interestingly, many other types of
network, including computer networks, biological neural networks, metabolic
networks~\autocite{jeong2000large}, and academic co-author networks, also have
the scale-free property.

Several properties of scale-free networks are relevant in epidemiology.  The
high-degree hub nodes are known as
\defn{superspreaders}~\autocite{kemper1980identification}, which have been
postulated to contribute in varying degree to the spread of diseases such as
\gls{HIV}~\autocite{stadler2013uncovering} and
\gls{SARS}~\autocite{shen2004superspreading}. Scale-free networks have no
epidemic threshold~\autocite{pastor2001epidemic}, meaning that diseases with
arbitrarily low transmissibility can persist at low levels indefinitely. This
is in contrast with homogeneously mixed populations, in which transmissibility
below the epidemic threshold would result in exponential decay in the number of
infected individuals and eventual extinction of the pathogen (Anderson \& May,
I think).

@@

There has been some contention of the idea that contact networks are
scale-free. \textcite{handcock2004likelihood} fit several stochastic models of
partner formation to empirical degree distributions derived from population
surveys of sexual behaviour. They found that a negative binomial distribution,
rather than a power law, was the best fit to five out of six datasets, although
the difference in goodness of fit was extremely small in four out of these
five. \textcite{bansal2007individual} found that an exponential distribution,
rather than a power law, was the best fit to degree distributions of six social
or sexual networks.

@@

The contact network underlying an epidemic constrains the shape of the
transmission network, which in turn determines the topology of the transmission
tree relating the infected hosts (\cref{fig:contactnet}). The initially
infected node who introduces the epidemic into the network, also known as the
\defn{index case}, becomes the root of the tree. Each time a transmission
occurs, the lineage corresponding to the donor host in the tree splits into
two, representing the recipient lineage and the continuation of the donor
lineage. \Cref{fig:contactnet} illustrates this correspondence. It's important
to note that, although the order and timing of transmissions determines the
tree topology uniquely, the converse does not hold. That is, for any given
topology, there are in general many transmission networks which would lead to
that topology. In other words, it impossible to distinguish who transmitted to
whom from a transmission tree alone~\autocite{bernard2007hiv}.

A number of studies have made progress in quantifying the relationship between
contact networks and transmission trees. \textcite{o2010contact} simulated
epidemics over networks with four types of degree distribution. They then
estimated the Bayesian skyride~\autocite{minin2008smooth} population size
trajectory in two ways: from the phylogeny, using \gls{MCMC}; and from the
incidence and prevalence trajectories, using the method developed by 
\textcite{volz2009phylodynamics}. They found that the concordance between
the two skyrides, as well as the relationship between the skyride and
prevalence curve, was qualitatively different for each degree distribution.
\citeauthor{leventhal2012inferring} simulated transmission trees over \gls{ER},
\gls{WS}, \gls{BA}, and full networks with fixed number of nodes and mean
degree. They calculated Sackin's index of the simulated trees while varying the
epidemic, network, and sampling parameters, and found that the relationship
between these parameters and Sackin's index varies considerably among the
different network models. In summary, studies in this group have demonstrated
that network structure profoundly influences tree shape, but have not attempted
to quantitatively infer network parameters from observed trees.
